---
title: "20220516_AtlasToSpatial_TransferLearning"
author: "Jacob Mitchell"
date: "5/16/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r}
library(ggplot2)
library(ggpubr)
library(CoGAPS)
library(plyr)
library(dplyr)
library(forcats)
library(monocle3)
library(Seurat)
library(projectR)
library(fgsea)
library(msigdbr)
library(dittoSeq)

# Increase memory limit for R on Windows
if (memory.limit() < 15000){
  memory.limit(size = 15000)
}

sessionInfo()
```

```{r}
# load in Luciane's Panin visium data
load("data/20220428_panin_epithelial/panin_epithelial.Rda")

# read in the CoGAPS results object and cds object subset to epithelial cells
cogaps <- readRDS("../PDAC_Atlas_Pipeline/CoGAPS_output/epiMat-e74511f6-8a1d-4929-ab42-1cb67c5a4fb6-result-8pattern.rds")
cds <- readRDS("../PDAC_Atlas_Pipeline/CoGAPS/cds_combined_epithelial_Peng&Steele_8_SHARED_GENES_QC_filtered_harmonized_preprocessed_aligned_manuscript_UMAP_learn_graph(6).rds")
```

### CoGAPS Pattern Marker Cells and Pattern Weight Assignments for PDAC Atlas

```{r}
# Assign cells as pattern markers based on the patternMarkers function
## Run the pattern marker statistic on the cells 
cellPatternMarker <- patternMarkers(cogaps, axis = 2)

## Add the pattern to the cell marker result
for(i in 1:length(cellPatternMarker$PatternMarkers)){
  ## Concert to a df
  cellPatternMarker$PatternMarkers[[i]] <- as.data.frame(cellPatternMarker$PatternMarkers[[i]])
  ## create column name
  colnames(cellPatternMarker$PatternMarkers[[i]]) <- "CellID"
  cellPatternMarker$PatternMarkers[[i]]$Pattern <- names(cellPatternMarker$PatternMarkers[i])
}

## Combined the tables
cellPatternMarkerDataFrame <- do.call("rbind", cellPatternMarker$PatternMarkers)
## Update rownames
rownames(cellPatternMarkerDataFrame) <- cellPatternMarkerDataFrame$CellID

## Add the pattern assignment to the cds
colData(cds)$AssignedPattern <- cellPatternMarkerDataFrame[colnames(cds), "Pattern"]
```

```{r}
# Incorporate pattern weights learned from CoGAPS in the cds metadata
# Extract the pattern matrix
patMat <- cogaps@sampleFactors
# Merge with the column data in the cds
cds@colData <- cbind(colData(cds), patMat[colnames(cds), ])
# data frame of the gene names and ensembl IDs of the cell features
cds_gene_names <- as.data.frame(cds@rowRanges@elementMetadata@listData[c("gene_ensembl_ID", "gene_short_name")])
```

```{r}
# create an additional classifier for cells that divides Epithelial_normal
# into normal cells from non-PDAC patients and normal_tumor_adjacent cells
# from patients with PDAC

cds@colData$Epithelial_Cell_Type <- paste0(as.character(cds@colData$TN),"_", 
                                    as.character(cds@colData$TN_assigned_cell_type_immune_broad))
cds@colData$Epithelial_Cell_Type <- gsub("T_Epithelial_cancer", "Cancer",
                                    gsub("T_Epithelial_unspecified", "Unspecified",
                                    gsub("T_Epithelial_normal", "Normal_Tumor_Adjacent",
                                    gsub("N_Epithelial_cancer", "Cancer",
                                    gsub("N_Epithelial_unspecified", "Unspecified",
                                    gsub("N_Epithelial_normal", "Normal", cds@colData$Epithelial_Cell_Type))))))
```

### Projection of PDAC Atlas Patterns to PanIN Visium Data

```{r}
# project CoGAPS pattern weights from PDAC atlas epithelial cells onto
# SCT scaled transcript matrix of panin visium epithelial cells
project_embed <- projectR(data = panin_epithelial@assays$SCT@scale.data, loadings = cogaps)

# add projected patterns to seurat object
pattern_weights <- as.data.frame(t(project_embed))

for(pattern in colnames(pattern_weights)) {
  panin_epithelial <- AddMetaData(object = panin_epithelial,
                                  metadata = pattern_weights[[pattern]],
                                  col.name = pattern)
}

```

### Creation of Results Directories

```{r}
# order tumor grade annotation from normal to low to high
panin_epithelial$Grade <- factor(panin_epithelial$Grade,
                                 levels = c("N", "LG", "HG"))
# Palette for grade colors
# yellow, orange, red
Grade_pal <- c("#EBF4A5", "#F2A93B", "#EA3F25")
# list of CoGAPS Patterns
patterns <- colnames(pattern_weights)
# list of Panin Images
panin_images <- Images(panin_epithelial)

# filepath to save results and figures
figurepath <- "results/20220516_figures"
if (!dir.exists(figurepath)){
  dir.create(figurepath, recursive = TRUE)
}
# Directory of Violin Plots
Vlnpath <- paste0(figurepath, "/Projected_Pattern_ViolinPlots")
if (!dir.exists(Vlnpath)){
  dir.create(Vlnpath, recursive = TRUE)
}
# Directory of Pattern Weights on Slides
PatternSpatialpath <- paste0(figurepath, "/Projected_Pattern_SpatialPlots")
if (!dir.exists(PatternSpatialpath)){
  dir.create(PatternSpatialpath, recursive = TRUE)
}
# Directory of Epithelial Grade on Slides
GradeSpatialpath <- paste0(figurepath, "/Grade_SpatialPlots")
if (!dir.exists(GradeSpatialpath)){
  dir.create(GradeSpatialpath, recursive = TRUE)
}
# Directory of Pattern Weights in the PDAC Atlas
PatternAtlaspath <- paste0(figurepath, "/Pattern_AtlasPlots")
if (!dir.exists(PatternAtlaspath)){
  dir.create(PatternAtlaspath, recursive = TRUE)
}
# TFF1 Expression Plots
TFF1path <- paste0(figurepath, "/TFF1_Expression")
if (!dir.exists(TFF1path)){
  dir.create(TFF1path, recursive = TRUE)
}
# Cell Types from PDAC Atlas
CellTypepath <- paste0(figurepath, "/Atlas_Cell_Types")
if (!dir.exists(CellTypepath)){
  dir.create(CellTypepath, recursive = TRUE)
}


# Save RDS of the visium data with projected pattern weights
saveRDS(panin_epithelial,
        file = paste0(figurepath, "/panin_epithelial_projected_cogaps_patterns.rds"))
```

### Custom Plotting Functions

```{r}
# functions for consistent plotting parameters

# Pattern Weight UMAP Plots for the PDAC Atlas
plotUMAP <- function(cds, color_cells_by, title, show = TRUE) {
  plot <- plot_cells(cds, color_cells_by = color_cells_by,
                     label_cell_groups = FALSE, show_trajectory_graph = FALSE) +
    theme(legend.position = "bottom") +
    ggtitle(title)
  if(show){
    print(plot)
  }
  return(plot)
}

# Violin Plots: Projected Pattern Weight in Each Tumor Grade
plotViolin <- function(seurat, feature, title, show = TRUE){
  violin <- VlnPlot(seurat, features = feature, group.by = "Grade",
                  cols = Grade_pal)
  # save minimum and maximum y axis limits for feature to ensure that space is left
  # for the ggpubr significance marks
  
  feat_range <- layer_scales(violin)$y$get_limits()
  feat_dif <- feat_range[2] - feat_range[1]
  y_min <- feat_range[1]
  y_max <- feat_range[2] + feat_dif/2.5
  
  plot <- violin +
    ylim(c(y_min, y_max)) +
    stat_compare_means(comparisons = list(
      c("N", "LG"),
      c("N", "HG"),
      c("LG", "HG"))) +
    ggtitle(title) + 
    ylab(paste0("projected ", feature, " weight")) +
    xlab("epithelial lesion grade")
  if(show){
    print(plot)
  }
  return(plot)
}

# Spatial Plots: Projected Pattern Weight in Spots on Slides
plotPatternSpatial <- function(seurat, feature, image, show = TRUE){
  plot <- SpatialFeaturePlot(seurat, features = feature, images = image,
                             pt.size.factor = 3)
  if(show){
    print(plot)
  }
  return(plot)
}

# Spatial Plots: Tumor Grade in Spots on Slides
plotGradeSpatial <- function(seurat, group, image, show = TRUE){
  plot <- SpatialDimPlot(seurat, group.by = group, images = image,
                         pt.size.factor = 3, cols = Grade_pal)
  if(show){
    print(plot)
  }
  return(plot)
}
```

### Plot Projected Pattern Weights

```{r}
for(p in patterns){
  vln_plot <- plotViolin(panin_epithelial, feature = p, title = NULL)
  ggsave(filename = paste0(Vlnpath, "/", p, "_Projected_Weights.pdf"),
         plot = vln_plot,
         width = unit(6, "in"),
         height = unit(6, "in"))
}
```

### Plot Projected Pattern Weights and Epithelial Cell Grades on image data

```{r}
for(i in panin_images){
  i_name <- gsub("\\.", "_", i)
  
  grade_plot <- plotGradeSpatial(panin_epithelial, group = "Grade", image = i) 
  ggsave(filename = paste0(GradeSpatialpath, "/", i_name, "_Epithelial_Grade.pdf"),
         plot = grade_plot,
         width = unit(10, "in"),
         height = unit(10, "in"))
  
  # nested loop for plotting each pattern's weight on each image
  for(p in patterns){
    pat_plot <- plotPatternSpatial(panin_epithelial, feature = p, image = i) +
      scale_fill_viridis_c(option = "magma") +
      theme(legend.position="right")
    
    ggsave(filename = paste0(PatternSpatialpath, "/", p, "_", i_name, "_Epithelial_Projection.pdf"),
           plot = pat_plot,
           width = unit(10, "in"),
           height = unit(10, "in"))
  }
}
```

### Plot UMAPs of Pattern Weights in PDAC Atlas

```{r}
for(p in patterns){
  umap_plot <- plotUMAP(cds = cds, color_cells_by = p, title = NULL) +
    scale_color_viridis_c(option = "magma") +
    labs(color=p)
  ggsave(filename = paste0(PatternAtlaspath, "/", p, "_Atlas_Weights.pdf"),
         plot = umap_plot,
         width = unit(10, "in"),
         height = unit(10, "in"))
}
```

### UMAP Plot of TFF1 Expression in Atlas

```{r}
TFF1_plot <- plot_cells(cds, genes = "TFF1",
             label_cell_groups = FALSE, show_trajectory_graph = FALSE) +
  theme(legend.position = "bottom") +
  scale_color_viridis_c()
print(TFF1_plot)
ggsave(filename = paste0(TFF1path, "/TFF1_Atlas_Expression.pdf"),
       plot = TFF1_plot,
       width = unit(10, "in"),
       height = unit(10, "in"))

# TFF1 Expression in Panin Visium
for(i in panin_images){
  i_name <- gsub("\\.", "_", i)
  TFF1_spatial_plot <- plotPatternSpatial(panin_epithelial, feature = "TFF1", image = i) +
    scale_fill_viridis_c() +
    theme(legend.position="right")
  print(TFF1_spatial_plot)
  ggsave(filename = paste0(TFF1path, "/TFF1_", i_name, "_Spatial_Expression.pdf"),
           plot = TFF1_spatial_plot,
           width = unit(10, "in"),
           height = unit(10, "in"))
}

# Violin Plot of TFF1 Expression in Panin Visium
TFF1_vln_plot <- plotViolin(panin_epithelial, feature = "TFF1", title = NULL) +
  ylab("TFF1 Expression")
ggsave(filename = paste0(TFF1path, "/TFF1_PanIN_Expression.pdf"),
       plot = TFF1_vln_plot,
       width = unit(6, "in"),
       height = unit(6, "in"))
```

### UMAP of Epithelial Cell Types in the PDAC Atlas

```{r}
epi_plot <- plotUMAP(cds, color_cells_by = "Epithelial_Cell_Type", title = NULL) +
  scale_color_manual(values = dittoColors()[c(3,2,5,7)])
print(epi_plot)
ggsave(filename = paste0(CellTypepath, "/Epithelial_Atlas_Cell_Types.pdf"),
       plot = epi_plot,
       width = unit(10, "in"),
       height = unit(10, "in"))
```

### UMAP of Epithelial Cell Types in ST Panin Data

```{r}
# get metadata
md <- panin_epithelial[[]]
# get embeddings
coords <- Embeddings(panin_epithelial[["umap"]])
# combine dataframes
md <- cbind(md, coords)

# plot
st_umap <- ggplot(md, aes(x = UMAP_1, y = UMAP_2, fill = Grade)) +
  scale_fill_manual(values = Grade_pal) +
  geom_point(shape = 21, size = 4, stroke = 0.1) + theme_classic()
print(st_umap)
ggsave(filename = paste0(CellTypepath, "/Epithelial_Spatial_Cell_Types.pdf"),
       plot = st_umap,
       width = unit(10, "in"),
       height = unit(10, "in"))
```

### Pattern Marker Overrepresentation and Plotting

```{r}
patternMarkerResults <- patternMarkers(cogaps, threshold = "cut", lp = NA, axis = 1)
names(patternMarkerResults$PatternMarkers) <- colnames(patternMarkerResults$PatternMarkerRanks)PMlist <- patternMarkerResults$PatternMarkers

patternmarks <- data.frame(lapply(patternMarkerResults$PatternMarkers, "length<-",
                                  max(lengths(patternMarkerResults$PatternMarkers))))

hallmark_df <- msigdbr(species = "Homo sapiens", category = "H")
hallmark_list <- hallmark_df %>% split(x = .$gene_symbol, f = .$gs_name)
for (pattern in names(PMlist)) {
  
  # Over-representation analysis of pattern markers ###########################
  suppressWarnings(
    result <- fora(pathways = hallmark_list,
                   genes = PMlist[[pattern]],
                   universe = Hs_genes$hgnc_symbol,
                   maxSize=2038)
  )
  
  # Append ratio of overlap/# of genes in hallmark (k/K)
  result$"k/K" <- result$overlap/result$size
  
  # Append log-transformed HB-adjusted q value (-10 * log(adjusted p value))
  result$"neg.log.q" <- (-10) * log10(result$padj)
  
  # Append shortened version of hallmark's name without "HALLMARK_"
  result$MsigDB_Hallmark <- substr(result$pathway, 10, nchar(result$pathway))
  
  # Reorder dataframe by ascending adjusted p value
  result <- mutate(result, MsigDB_Hallmark=fct_reorder(MsigDB_Hallmark, - padj))
  
  # rearrange columns to move overlap genes to the end and convert to vector for
  # compatibility with saving as csv
  result <- relocate(result, "overlapGenes", .after = "MsigDB_Hallmark")
  result <- relocate(result, "neg.log.q", .after = "padj")
  result <- result %>% mutate(overlapGenes = sapply(overlapGenes, toString))
  
  # save pattern marker overlaps as a csv file
  write.csv(result,
            file = paste0("results/MsigDB_ORA/", pattern, "_MsigDB_OverRepresentationAnalysis.csv"))
  
  # for plotting, limit the results to significant over-representation
  result <- result[result$padj < 0.05,]
  
  #plot and save the waterfall plot of ORA p-values
  plot <- ggplot(result, aes_string(y = "neg.log.q", x = "MsigDB_Hallmark", fill = "MsigDB_Hallmark")) +
    ## Specifies barplot
    geom_col() +
    ## Rename y axis
    ylab("-10*log10(FDR q-value)") + 
    ## Flips the coordinates
    coord_flip() +
    ## Makes the background white
    theme_minimal() +
    ## Add title
    ggtitle(paste0(pattern, ": Overrepresented MsigDB Hallmarks")) +
    ## This creates the dotted line at .05 value 
    geom_hline(yintercept=c(13.0103), linetype="dotted") + # Add veritcle line to show significances
    ## Adds the q values
    geom_text(aes(label=format(signif(padj, 4))), hjust = -.04) +
    ## Removes legend
    theme(legend.position = "none") +
    ## specifies limits 
    ylim(0, ceiling(max(result$"neg.log.q")) + (max(result$"neg.log.q")/4))
  ggsave(paste0("results/figures/Pattern_MsigDB_ORA/", pattern, "_ORA.pdf"),
         plot = plot,
         width = unit(10, "in"),
         height = unit(6, "in"),
         device = "pdf")
}
```

